White Dwarf Mergers on Adaptive Meshes\\ III. Mergers

### 2016-09-12 papers/wdmerger_III/paper.tex
Describe the inertial frame variable scheme

To implement this method in \castro, we desired an approach that was flexible
and would leave as little imprint on the code structure as possible.
To this end, what we have done is to add three new ``hybrid'' momentum state
variables, corresponding to the radial momentum, angular momentum, and the linear
momentum component that is perpendicular to both. Whenever we update the normal
momentum state variables with hydrodynamics fluxes, we add the analogous fluxes
to the hybrid momenta -- that is, using the edge state values determined by the
final multi-dimensional Riemann solve, we construct the value of the hybrid
momenta on that zone edge, and then allow it to be transported as usual by the
advective velocity $\mathbf{u}$. When we update the normal momenta with an
external forcing $\mathbf{F}$ such as gravity, with $x$ and $y$
components $F_x$ and $F_y$, we apply the update to the hybrid momenta as well:
\begin{align}
  \left.\frac{\partial(s_R)}{\partial t}\right|_{F} &= -F_x \frac{x}{R} - F_y\frac{y}{R} \label{eq:force-radial}\\
  \left.\frac{\partial(\ell_z)}{\partial t}\right|_{F} &= F_x\, y - F_y\, x.\label{eq:force-angular}
  \end{align}

If in addition to evolving these variables we want to use them in place of the
standard linear momenta for determining the hydrodynamic flow, only one change
is needed for the standard timestep advance scheme: after source terms are
applied to the state, we recompute the normal momenta so that they are fully
consistent with the hybrid momenta. In practice for a \castro\ advance this
synchronization of the linear and hybrid momenta occurs twice: once after the
old-time source terms and the hydrodynamic update has been applied (these all
use time-level $n$ data to construct the update, which is expected to already have
synchronized linear and hybrid momenta), and once after new-time source
terms have been applied. The use of adaptive mesh refinement implies sychronization
in two other places: after a reflux that syncs up the fluxes between coarse and
fine grids; and, after interpolation has been performed to obtain fine grid data
from a coarser grid (since the spatially interpolated data may not respect the
consistency relationship between the two momenta). In this way, the linear momenta
are essentially placeholder variables that always reflect what the current radial
and angular momenta are, but it is quite useful to express the scheme in this way
because most of the \castro\ infrastructure is already set up to use the linear
momentum state variables for calculating source terms, so the hybrid scheme requires
minimal changes to the code architecture. For example, when we compute
the primitive variables from the conservative variables in preparation for an advective
hydrodynamic update, we will still be getting them from the linear momenta,
but these linaer momenta are consistent with the conserved angular momentum,
so we get the same result as if we had modified the code to explicitly compute
the primitive variables as a function of the hybrid momenta. Consequently the
scheme presented here adds the ability to explicitly conserve angular momentum
in a simulation without a significant footprint on the simulation software.

>>>>

To implement this method in \castro, we desired an approach that was flexible
and would leave as little imprint on the code structure as possible.
To this end, what we have done is to add three new ``hybrid'' momentum state
variables, corresponding to the radial momentum, angular momentum, and the linear
momentum component that is perpendicular to both. Whenever we update the normal
momentum state variables with hydrodynamics fluxes, we add the analogous fluxes
to the hybrid momenta --- that is, using the edge state values determined by the
final multi-dimensional Riemann solve, we construct the value of the hybrid
momenta on that zone edge, and then allow it to be transported as usual by the
advective velocity $\mathbf{u}$. When we update the normal momenta with an
external forcing $\mathbf{F}$ such as gravity, with $x$ and $y$
components $F_x$ and $F_y$, we apply the update to the hybrid momenta as well:
\begin{align}
  \left.\frac{\partial(s_R)}{\partial t}\right|_{F} &= -F_x \frac{x}{R} - F_y\frac{y}{R} \label{eq:force-radial}\\
  \left.\frac{\partial(\ell_z)}{\partial t}\right|_{F} &= F_x\, y - F_y\, x.\label{eq:force-angular}
  \end{align}

If in addition to evolving these variables we want to use them in place of the
standard linear momenta for determining the hydrodynamic flow, only one change
is needed for the standard timestep advance scheme: after source terms are
applied to the state, we recompute the normal momenta so that they are fully
consistent with the hybrid momenta. In practice for a \castro\ advance this
synchronization of the linear and hybrid momenta occurs twice: once after the
old-time source terms and the hydrodynamic update has been applied (these all
use time-level $n$ data to construct the update, which is expected to already have
synchronized linear and hybrid momenta), and once after new-time source
terms have been applied. The use of adaptive mesh refinement implies sychronization
in two other places: after a reflux that syncs up the fluxes between coarse and
fine grids; and, after interpolation has been performed to obtain fine grid data
from a coarser grid (since the spatially interpolated data may not respect the
consistency relationship between the two momenta). In this way, the linear momenta
are essentially placeholder variables that always reflect what the current radial
and angular momenta are, but it is quite useful to express the scheme in this way
because most of the \castro\ infrastructure is already set up to use the linear
momentum state variables for calculating source terms, so the hybrid scheme requires
minimal changes to the code architecture. For example, when we compute
the primitive variables from the conservative variables in preparation for an advective
hydrodynamic update, we will still be getting them from the linear momenta,
but these linear momenta are consistent with the conserved angular momentum,
so we get the same result as if we had modified the code to explicitly compute
the primitive variables as a function of the hybrid momenta. Consequently the
scheme presented here adds the ability to explicitly conserve angular momentum
in a simulation without a significant footprint on the simulation software.

\subsection{Rotation Revisited}\label{sec:rotation}

Another significant change we have made in \castro\ since \citet{wdmergerI} deals
with how rotation is applied as a source term. We noted in that paper that rotation
can be used to reduce advection errors in a binary system due to the fact that
material is moving less on the grid. However, we have also observed that rotation
source terms introduce the opportunity for other numerical problems to arise. In
particular, the most significant problem for us is that conservation of linear and
angular momentum is harder to accomplish with the standard rotation source terms.
Largely this is due to the fact that the rotation force is not expressed in a numerically
conservative fashion; the forces are evaluated at cell centers and do not inherently
have the telescoping property that allows for conservation of momentum in the case
of hydrodynamic fluxes. While gravity has the property that in practice momentum
conservation errors due to the gravitational force are typically quite small due to
an underlying telescoping behavior in the gravitational force (see the discussion
in Section 2.3.1 of \cite{wdmergerI}), this is harder to obtain for rotation (this is
easy to see when considering the Coriolis force, as the standard Coriolis forcing
involves cell-centered velocities that cannot trivially be evaluated in a way that
leads to numerical cancellation --- though see \citet{audusse:2009} for a method that
utilizes the edge-centered hydrodynamic fluxes to this end). A way around this issue
is to evolve variables measured in the \textit{inertial} frame, even when the reference
frame used for the advection is rotating. The benefits of this were observed by
\citet{kley:1998} in the context of protostellar accretion disk simulations (and the
method has since been used in a number of codes in that community, e.g. \citet{NIRVANA}
and \citet{FARGO3D}), and also by \citet{call:2010}, whose method was adopted by
\citet{byerly:2014} in the context of binary star simulations. The benefit of using
inertial frame variables in this context is that the rotational terms vanish as an explicit
source term to the angular momentum. \citet{kley:1998} points out that the rotational update
to the $z$-component of the angular momentum measured in the rotating frame (which we
temporarily call $\ell^\prime$) is:
\begin{equation}
  \left.\frac{\partial \ell}{\partial t}\right|^{\text{rotation}} = -2\rho\, v_r\, \omega\, R.\label{eq:rotating-angular-momentum}
\end{equation}
(Note that in \castro\ we construct all source terms, including rotation, for the linear
momenta and then apply the source term to the hybrid momenta using \autoref{eq:force-radial} and
\autoref{eq:force-angular}; \autoref{eq:rotating-angular-momentum} is automatically obtained this way.)
The term on the right-hand-side vanishes if instead we evolve the inertial frame $\ell$, and there
is no source term for the radial momentum either. If one is evolving the linear momenta measured
in the inertial frame, \cite{byerly:2014} give the form of the update:
\begin{equation}
  \left.\frac{\partial (\rho \mathbf{u})}{\partial t}\right|^{\text{rotation}} = -{\bm{\omega}} \times (\rho \mathbf{u})
\end{equation}
This is equal to one-half of the Coriolis force (with the centrifugal force not used; see
\cite{call:2010} for a further discussion of the meaning of this expression),
and is needed because the linear momentum components change when advected over a
rotating cylindrical grid. (Since the form of the update is similar to the Coriolis force,
we can still use the implicit update technique from Section 2.4 of \cite{wdmergerI} when
evaluating the new-time source term at time-level $n+1$.)

The ability to evolve state variables measured in the inertial frame when doing a
rotating frame simulation has been added as a runtime parameter in \castro. Note that
other than evolving inertial frame variables, other aspects of the rotating frame are
preserved --- for example, binary stars in a stable circular orbit will remain in the same
location on the grid as where they started, just as in a standard rotating frame simulation,
and the difference is mainly that the simulation velocity associated with those stationary
stars will read as the equivalent velocity of the star in the inertial frame. The only
meaningful wrinkle this introduces is that the hydrodynamic update needs to use the
rotating frame velocities, not the inertial frame velocities. So in the conversion from
conserved variables $\mathbf{U}$ to primitive variables $\mathbf{q}$, the latter being
the ones used by the hydrodynamics module to construct an advective update, we subtract
the rotating frame velocity from the conserved variables when constructing $\mathbf{q}$.
Then the rest of the hydrodynamics update proceeds as normal with rotating frame velocities.
As observed by \citet{byerly:2014}, this is acceptable because the divergence of the
rotational velocity field is zero, so it can legitimately be analytically subtracted
from the inertial velocity field. The payoff to the scheme is that we get the better
numerical behavior associated with the rotating frame velocities, as they are smaller
in magnitude with respect to the grid.

\subsection{Miscellaneous Other Changes}
\label{sec:hydro-other-changes}

- Density flux limiter
